---
title: "BGS PCA"
author: "Kelly Murman"
date: "`r Sys.Date()`"
output: html_document
---

```{r load packages, include = F}

library(FactoMineR)
library(factoextra)
library(vegan)
library(tidyverse)
library(ggrepel)
library(corrplot)

```

```{r load data, include = F}

backsamps <- read.csv("BGS SITE MATRIX.csv")

agebacksamps <- read.csv("BGS SITE MATRIX v2.csv")

```

#PCA of environmental background samples using FactoMineR and factoextra

#Environmental background samples were collected from terrestrial and aquatic sources. Leaf litter samples were collected from the stream banks and leaf packs were placed in the streams for four weeks prior to the start of salamander sampling to sample for aquatic invertebrates.

#Question: Is there a difference in the types of invertebrates present in leaf litter samples and leaf pack samples, i.e., can these samples be used to classify invertebrates as aquatic or terrestrial?

```{r PCA}

backsamps.pca <- PCA(backsamps, quali.sup = 1:4, graph = T)

plotellipses(backsamps.pca)

get_eig(backsamps.pca)

fviz_screeplot(backsamps.pca, addlabels = TRUE, ylim = c(0,15))

# extract the results for variables
var <- get_pca_var(backsamps.pca)
var

# coordinates of variables
head(var$coord)
varcoord <- head(var$coord, 84)

# contribution of variables
varcontrib <- head(var$contrib, 84)

# quality of representation = cos2 of variables
varcos2 <- head(var$cos2, 84)

corrplot(var$cos2, is.corr = FALSE)

fviz_cos2(backsamps.pca, choice = "var", axes = 1:2)

# original archery target
fviz_pca_var(backsamps.pca, col.var = "cos2", gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE)

# graph of variables: default plot
fviz_pca_var(backsamps.pca, col.var = "black")


# control variable colors using their contributions
# *** this is the archery target plot, how to remove labels? I added the "select.ind" argument, but it is still extremely busy:
fviz_pca_var(backsamps.pca, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             select.var = list(contrib = 10), repel = TRUE) ##avoid text overlapping


# contributions of variables to PC1
fviz_contrib(backsamps.pca, choice = "var", axes = 1, top = 10)

fviz_contrib(backsamps.pca, choice = "var", axes = 1, top = 84)

# contributions of variables to PC2
fviz_contrib(backsamps.pca, choice = "var", axes = 2, top = 10)

# extract the results for individuals
ind <- get_pca_ind(backsamps.pca)
ind

# coordinates of individuals
head(ind$coord)

# graph of individuals
# 1. use repel = TRUE to avoid overplotting
# 2. control automatically the color of individuals using the cos2
  # cos2 = the quality of the individuals on the factor map
  # use points only
# 3. use gradient color
fviz_pca_ind(backsamps.pca, col.ind = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE)

# biplot of individuals and variables
fviz_pca_biplot(backsamps.pca, repel = TRUE)

# color coded graph of individuals
fviz_pca_ind(backsamps.pca, label = "none", habillage = as.factor(backsamps$samp.type), palette = c("#00AFBB", "#E7B800", "#FC4E07"), addEllipses = TRUE)

fviz_pca_ind(backsamps.pca, label = "none", habillage = as.factor(backsamps$season), palette = c("#00AFBB", "#E7B800", "#FC4E07"), addEllipses = TRUE)

fviz_pca_ind(backsamps.pca, label = "none", habillage = as.factor(backsamps$site), palette = c("#00AFBB", "#E7B800", "#FC4E07", "#032d70"), addEllipses = TRUE)

fviz_pca(backsamps.pca, label = "none", geom.ind = "point", geom.var = "text")


# trying to overlay the graph of individuals ~ sample type with the graph of variables (archery target) - should be able to do this with the biplot function (biplot of individuals and variables)

fviz_pca_biplot(backsamps.pca, label = "var", select.var = list(contrib = 10), habillage = as.factor(backsamps$samp.type), addEllipses = TRUE, repel = TRUE, palette = c("#00AFBB", "#E7B800"))


```


#Analysis of Similarities (ANOSIM)

```{r ANOSIM}

com <- backsamps[,5:ncol(backsamps)]

m_com <- as.matrix(com)

# getting a new error message when I try to run this now...

anosim(x = m_com, grouping = backsamps$samp.type, permutations = 9999, distance = "bray")

anosim(x = m_com, grouping = backsamps$site, permutations = 9999, distance = "bray")

anosim(x = m_com, grouping = backsamps$season, permutations = 9999, distance = "bray")

anosim(x = m_com, grouping = backsamps$samp.point, permutations = 9999, distance = "bray")

```


#PCA with taxa separated by age class.


```{r PCA with taxa by age class}

agebacksamps <- agebacksamps[, -5]

agebacksamps.pca <- PCA(agebacksamps, quali.sup = 1:4, graph = T)

plotellipses(agebacksamps.pca)

get_eig(agebacksamps.pca)

fviz_screeplot(agebacksamps.pca, addlabels = TRUE, ylim = c(0,15))

## extract the results for variables
agevar <- get_pca_var(agebacksamps.pca)
agevar

## coordinates of variables
head(agevar$coord)
ageeigenvalues <- head(agevar$coord, 70)
ageeigenvalues

## contribution of variables
head(agevar$contrib, 70)

## graph of variables: default plot
fviz_pca_var(agebacksamps.pca, col.var = "black")

## control variable colors using their contributions
fviz_pca_var(agebacksamps.pca, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE) ##avoid text overlapping

## contributions of variables to PC1
fviz_contrib(agebacksamps.pca, choice = "var", axes = 1, top = 10)

## contributions of variables to PC2
fviz_contrib(agebacksamps.pca, choice = "var", axes = 2, top = 10)

## extract the results for individuals
ageind <- get_pca_ind(agebacksamps.pca)
ageind

## coordinates of individuals
head(ageind$coord)

## graph of individuals
## 1. use repel = TRUE to avoid overplotting
## 2. control automatically the color of individuals using the cos2
  # cos2 = the quality of the individuals on the factor map
  # use points only
## 3. use gradient color
fviz_pca_ind(agebacksamps.pca, col.ind = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE)

## biplot of individuals and variables
fviz_pca_biplot(agebacksamps.pca, repel = TRUE)

## color coded graph of individuals
fviz_pca_ind(agebacksamps.pca, label = "none", habillage = as.factor(agebacksamps$Site), palette = c("#00AFBB", "#E7B800", "#FC4E07", "#032d70"), addEllipses = TRUE)

fviz_pca_ind(agebacksamps.pca, label = "none", habillage = as.factor(agebacksamps$Season), palette = c("#00AFBB", "#E7B800", "#FC4E07", "#032d70"), addEllipses = TRUE)

fviz_pca_ind(agebacksamps.pca, label = "none", habillage = as.factor(agebacksamps$Samp.Type), palette = c("#00AFBB", "#E7B800", "#FC4E07", "#032d70"), addEllipses = TRUE)

fviz_pca_ind(agebacksamps.pca, label = "none", habillage = as.factor(agebacksamps$Age.Class), palette = c("#00AFBB", "#E7B800", "#FC4E07", "#032d70"), addEllipses = TRUE)

```

# Analysis of Similarities with age class as a factor

```{r ANOSIM w age class}
agecom <- agebacksamps[,5:ncol(agebacksamps)]

m_agecom <- as.matrix(agecom)

anosim(x = m_com, grouping = agebacksamps$Site, permutations = 999, distance = "bray")

anosim(x = m_com, grouping = agebacksamps$Season, permutations = 999, distance = "bray")

anosim(x = m_com, grouping = agebacksamps$Samp.Type, permutations = 999, distance = "bray")

anosim(x = m_com, grouping = agebacksamps$Age.Class, permutations = 999, distance = "bray")



```

## Stef's code using vegan for PCA

```{r vegan pca code, include = F}

pca_1 <- rda(backsamps[,5:ncol(backsamps)])
summary(pca_1)

taxa <- scores(pca_1, display = "species", scaling = 1)
site <- scores(pca_1, display = "sites", scaling = 1)

site_scores <- as.data.frame(site)
taxa <- as.data.frame(taxa)

site_scores$site <- backsamps$site
site_scores$season <- backsamps$season
site_scores$type <- backsamps$samp.type
site_scores$point <- backsamps$samp.point

backsamps$PC1 <- site_scores[,1]
backsamps$PC2  <- site_scores[,2]

hull <- backsamps %>% group_by(season) %>%
slice(chull(PC1, PC2))


pal <- c("#920000", "#db6d00","#009292")


# plot_sites <-
#   ggplot(env_var, aes(x = RDA1, y = RDA2)) +
#   #first site points, colored by a factor here - survey number
#   geom_point(aes(color = as_factor(env_var$survey_number), size = 3)) +
#   # I had issues with coloring - I tried to use the palette above (pal) but it wasn't working out for me since I wanted to color different elements on this figure so instead I manually filled in the colors. you should be able to set a palette and then call that palette name for fill or color
#   scale_color_manual(values = c("#920000", "#009292", "#db6d00")) +
#   labs(y = paste0("RDA1 (", round(abun_rda_final$CCA$eig[1] *10, 2), "%)"),
#        x = paste0("RDA2 (", round(abun_rda_final$CCA$eig[2] * 10, 2), "%)"),
#        title = "RDA Biplot - Site Scores") +
#   #then polygons
#   geom_polygon(data = hull, alpha = 0.01,
#                  aes(colour = as_factor(survey_number),
#                      fill = as_factor(survey_number)), show.legend = F) +
#   #then "species" points
#   geom_point(data = species_scores, aes(x = RDA2, y = RDA1,
#                                         fill = "black", size = 3), show.legend = F) +
#   geom_label_repel(data = species_scores, aes(x = RDA2, y = RDA1, label = rownames(species_scores)),
#                    size = 5, direction = "both", max.overlaps = 20, show.legend = F, label.size = NA) +
#   #below is for arrows - you don't need this
#   geom_segment(data = as.data.frame(abun_rda_final$CCA$biplot),
#                aes(x = 0, y = 0, xend = RDA2 * 1, yend = RDA1 * 1),
#                arrow = arrow(length = unit(0.5, "cm")), size = 2, color = "#004949", show.legend = F) +
#   # labeling the arrows - not needed
#   geom_label(data = as.data.frame(abun_rda_final$CCA$biplot),
#              aes(x = RDA2 * .8, y = RDA1 * .8, label = c("Temperature", "SLF Fitted")),
#              color = "#004949", vjust = "bottom" , hjust = 0.8, show.legend = F, label.size = NA) +
#   #this theme just makes it look simple and clean - look up other themes! such as theme_bw()
#    theme_minimal()   +
#   #labeling the legend
#   labs(colour = "Survey Number") +
#   #changing text side of title and in figure text
#   theme(legend.title = element_text(size = 14),
#         legend.text = element_text(size = 12)) +
#   #this removes the legend but may be a bit wonky at times
#   guides(size= "none")



plot_sites <- 
  ggplot(site_scores, aes(x = PC1, y = PC2)) +
  geom_point(aes(color = season, size = 1)) +
  geom_label_repel(aes(x = PC1, y = PC2, label = site_scores$site), 
                   size = 5, direction = "both", max.overlaps = 20, show.legend = F, label.size = NA) 
  # geom_polygon(data = hull, alpha = 1,
  #                aes(colour = as_factor(season),
  #                    fill = as_factor(season)), show.legend = F) +
  # #then "species" points
  # geom_point(data = taxa, aes(x = PC1, y = PC2, 
  #                                       fill = "black", size = 1), show.legend = F) +
  #  theme_minimal()   +
  # #labeling the legend
  # labs(colour = "seasonr") + 
  # #changing text side of title and in figure text
  # theme(legend.title = element_text(size = 14),
  #       legend.text = element_text(size = 12)) +
  # #this removes the legend but may be a bit wonky at times
  # guides(size= "none")  



plot_sites


```